---
title: "p8105_hw2_ejc2222"
author: "Emanuel Clemente"
date: "2026-10-08"
output: github_document
---

```{r setup, message = FALSE}
library(tidyverse)
```

## Problem 1

### Reading and cleaning the data

```{r}
transit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |> 
  janitor::clean_names() |> 
  select(line, station_name, station_latitude, station_longitude,
         route1:route11, entry, vending, entrance_type, ada) |> 
  mutate(entry = if_else(entry == "YES", TRUE, FALSE))
```

This dataset contains information on each entrance and exit for each subway station in New York City. The variables retained are line, station name, station latitude/longitude, routes served (`route1` through `route11`), entry, vending, entrance type, and ADA compliance. I read in the data with `read_csv()`, cleaned the variable names with `janitor::clean_names()`, kept the specified variables with `select()`, and used `mutate()` to convert `entry` from "YES"/"NO" to a logical variable (`TRUE`/`FALSE`). The resulting dataset has `r nrow(transit_df)` rows and `r ncol(transit_df)` columns. The NAs in the route columns reflect stations that serve fewer than 11 routes, not missing data.

### Questions

```{r}
n_stations = 
  transit_df |> 
  distinct(line, station_name) |> 
  nrow()

n_ada = 
  transit_df |> 
  filter(ada) |> 
  distinct(line, station_name) |> 
  nrow()

prop_no_vending = 
  transit_df |> 
  filter(vending == "NO") |> 
  pull(entry) |> 
  mean()
```

There are `r n_stations` distinct stations, `r n_ada` of which are ADA compliant. Approximately `r round(prop_no_vending * 100, 1)`% of station entrances/exits without vending allow entrance.

### Reformatting route number and route name

To make route number and route name distinct variables, I converted `route8` through `route11` to character so all route columns share the same type, then used `pivot_longer()` to collapse the 11 route columns into `route_number` and `route_name`, dropping empty route slots.

```{r}
transit_long_df = 
  transit_df |> 
  mutate(across(route8:route11, as.character)) |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    names_prefix = "route",
    values_drop_na = TRUE
  )
```

```{r}
n_a_stations = 
  transit_long_df |> 
  filter(route_name == "A") |> 
  distinct(line, station_name) |> 
  nrow()

n_a_ada = 
  transit_long_df |> 
  filter(route_name == "A", ada) |> 
  distinct(line, station_name) |> 
  nrow()
```

`r n_a_stations` distinct stations serve the A train, and `r n_a_ada` of those are ADA compliant.